Physcraper: Body

### 2020-07-01 2-body.Rmd
finished ejm comments

Public biological databases provide an amazing resource for evolutionary estimation. By linking molecular data, available from databases such as the GenBank [@benson2000genbank; @wheeler2000database], to alignments and phylogenies, available in
the TreeBASE repository [@piel2009treebase] and the Open Tree of Life datastore [@mctavish2015phylesystem], we can place new biological data in an evolutionary context.
However, connecting new sequence data with existing estimates of evolutionary relationships represents a challenge.

@@

A way to incorporate the best of two worlds (massive amounts of newly released molecular data AND fine-grained curation from human experts) is be to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. This expert-curated alignments can be continuously enriched and updated by incorporating newly released data from public molecular databases.

@@

The OpenTree tree repository [phylesystem; @mctavish2015phylesystem] automatically incorporates phylogenies uploaded into TreeBASE, and stores metadata linking the tree to its corresponding alignment repository in TreeBASE. However, if there are multiple alignments, TreeBASE does not indicate how they were used to generate the tree. This provides a loose means of linking the tree with the exact alignment that generated it.

Often linking data in an original alignment with its corresponding phylogeny has to be done by a human curator.
Moreover, different data repositories follow different systems for taxon and study identification, posing a real challenge to automatically link data from across databases that belong to the same taxon and study.
OpenTree's metadata system incorporates taxon identifiers from a variety of taxonomies and repositories, including the NCBI taxonomy, GBIF, etc., providing a way to automatically link data from different databases.

Physcraper is a python pipeline that relies on the OpenTree metadata system to connect databases through taxon identification numbers.
It's main functionality is to connect phylogenies stored in the OpenTree phylesystem, with alignments from TreeBASE and newly released DNA data from GenBank, in order to update previously known phylogenetic relationships in a continuous manner.
Because of its design, it allows taking advantage of the many resources provided by the OpenTree.
For example, it allows automatizing and standardizing the comparison of phylogenetic hypotheses with currently known relationships.

@@

- In order to take advantage of the OpenTree tools, the input tree needs to be stored in the
  OpenTree phylesystem [github.com/opentreeoflife/phylesystem](https://github.com/opentreeoflife/phylesystem).
  The main
  reason for this is that trees in phylesystem have a set of user defined characteristics
  that are essential for automatizing the phylogeny update process. The most relevant of these being the standardization of taxonomic names and the definition of ingroup and outgroup. Outgroup and ingroup taxa in the original tree are identified and tagged. This allows to automatically set the root for the updated tree on the final steps of the pipeline.
  Currently, only trees connected to
  a published study can be stored in the phylesystem (although there are plans to
  allow storage of unpublished trees).
  A user can choose from among the `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)). If the published tree you are interested in updating is not in there, you can upload it via OpenTree's curator tool ([www.opentreeoflife.org/curator](https://tree.opentreeoflife.org/curator).
- The alignment should be a gene alignment that was used to generate the tree. The original
  alignments are usually stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or the journal were the tree was originally published.
  If the alignment is stored in TreeBase, `physcraper` can download it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- A taxon name matching step is performed to verify that all taxon names on the tips
  of the tree are in the DNA character matrix and vice versa.
- A ".csv" file with the summary of taxon name matching is produced for the user.
- Unmatched taxon names are dropped from both the tree and alignment.
  Technically, just one matching name is needed to perform the searches. Please, see next section.
- A ".tre" file and a ".fas" file containing only the matched taxa are generated and saved in the `inputs` folder to be used in the following steps.

@@

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local-database sequence searches.
  For remote-database searches, we modified the BioPython [@cock2009biopython] BLAST function in the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternate BLAST URL. This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST without NCBI's required wait times.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accession numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the physcraper folder, allowing for secondary analyses to run significantly faster.
- Matched sequences will be discarded if the fall below a default e-value of 0.00001, and outside a default minimum and maximum length of 80% and 120%, respectively, of the average length of sequences in the checked alignment (gaps dropped).
These parameters can be configured for each run.
  This filtering guarantees that whole genome sequences are not included, as they
  All accepted sequences are assigned an internal identifier, and are further filtered.

- New sequences that are identical to existing sequences, or to subsets of an existing sequence are discarded, unless they reperesent a different taxon than the existing sequence.
Longer sequences belonging to the
  same taxon as the original sequence will be considered further for analysis.
- Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform cycles of sequence similarity search iteratively, by blasting the newly found sequences until no new sequences are found. By default only one BLAST cycle is performed and only sequences in the checked alignment are blasted.
- Accepted sequences are then downloaded in full, and stored as a local database in a directory that is globally accessible (default to physcraper/taxonomy), so they are accessible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

- The software MUSCLE [@edgar2004muscle] is used by default to perform sequence alignments.
- First, all new sequences are aligned using default MUSCLE options.
- Then, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
- The final alignment is not further processed automatically. So, we encourage users
  to check it either by eye and perform manual refinement or using any of the many
  tools for alignment processing, to eliminate columns with no information.
- Users may also use physcraper only to gather sequence matches, and apply their own preferred alignemnt and phylogenetic inference tools.

@@

- A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML [@stamatakis2014raxml], using default settings such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with default method. Currently only the number of bootsrap replicates can be modified by the user.
- The original tree is used as starting tree for the ML searches. It can also be set as a full topological constraint or not used at all, depending on the goals of the user.
- Bootstrap results are summarized with Dendropy [@sukumaran2010dendropy].
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by physcraper are defined by a taxon name space, capturing the NCBI accession information, as well as the taxon identifiers, allowing the user perform comparisons and conflict analyses.
- Robinson Foulds weighted and unweighted metrics between the tips in the input tree and those tips in the updated tree are calculated with Dendropy functions [@sukumaran2010dendropy].
- Finally a conflict analysis is performed. This is basically a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
- For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined.
- A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroups so trees are accurately rooted.

@@

We will illustrate the utility of physcraper in here with two use-case scenarios. One in which the user is interested in a particular group. Another one in which the user is interested in a particular phylogeny.
A tutorial as well as illustrated examples of commands for every step needed to perform a physcraper analysis are available elsewhere.

@@

A student is interested in the genus *Ilex*, the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.
The genus encompasses between 400-600 living species. A review of literature reveals that there are three published phylogenetic trees showing relationships within the hollies.
The first one has been made available in TreeBASE as well as in the OpenTree phylesystem and is part of the synthetic tree. It samples 48 species.
The second tree has not been made available anywhere, not even in the supplementary data of the original publication.
The most recent one has been made available in the OpenTree Phylesystem and in the DRYAD repository. It is the best sampled yet, with 200 species. However, it has not been added to the syntehtic tree yet.
This also makes it a perfect case to test the basic functionalities of physcraper: we know that the sequences of the most recently published tree have been made available on the GenBank database. Hence, we expect that updating the oldest tree will produce something very similar to the newest tree.

@@

All these tools are key efforts for advancing towards reproducibility in phylogenetics,
a field that has relied on processes which are somewhat artisanal.
In here, we highlight the potential of taking advantage of this careful curation work in previous phylogenetic estimates. By taking sources of information available from data repositories and present a method to link data from different repositories, while leveraging on the knowledge and intuition of the expert
community to build up our phylogenetic knowledge, we can use not only data accumulated in
molecular data repositories, but phylogenetic knowledge accumulated in phylogenetic tree repositories.
While not generating full phylogenies *de novo*, physcraper is still capable of generating new phylogenetic knowledge.
It can also be combined with data from repositories other than molecular data. For example geographic locations (GBIF), fossils (PBDB), etc.

>>>>

Public biological databases provide an amazing resource for evolutionary estimation. By linking molecular data, available from databases such as the GenBank [@benson2000genbank; @wheeler2000database], to alignments and phylogenies, available in
the TreeBASE repository [@piel2009treebase] and the Open Tree of Life (OpenTree) datastore [@mctavish2015phylesystem], we can place new biological data in an evolutionary context.
However, connecting new sequence data with existing estimates of evolutionary relationships represents a challenge.

@@

A way to incorporate the best of two worlds (massive amounts of newly released molecular data AND fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. This expert-curated alignments can be continuously enriched and updated by incorporating newly released data from public molecular databases.

@@

The Phylesystem (OpenTree's datastore) [@mctavish2015phylesystem] automatically incorporates phylogenies from TreeBASE, and saves metadata linking the original tree to its corresponding alignment repository in TreeBASE. If there are multiple alignments, TreeBASE does not always indicate how they were used to generate the tree. This provides a loose means of linking the tree with the exact alignment that generated it.

Often, linking data in an original alignment with its corresponding phylogeny has to be done by a human curator.
Moreover, different data repositories follow different systems for taxon and study identification, posing a real challenge to automatically link data from across databases that belong to the same taxon and study.
OpenTree's metadata system incorporates taxon identifiers from a variety of taxonomies and repositories, including the NCBI taxonomy, GBIF, etc., providing a way to automatically link data from different databases.

Physcraper is a Python encoded pipeline that relies on the OpenTree metadata system to connect databases through taxon identification numbers.
It's main functionality is to connect phylogenies stored in the OpenTree phylesystem, with alignments from TreeBASE and newly released DNA data from GenBank, in order to update previously known phylogenetic relationships in a continuous manner.
EXPLIAN WHY PYTHON IS BEST FOR THIS.
Because of its design, it allows taking advantage of the many resources provided by the OpenTree.
For example, it allows automatizing and standardizing the comparison of phylogenetic hypotheses with currently known relationships.

@@

- In order to take advantage of the OpenTree tools, the input tree needs to be stored in the
  OpenTree phylesystem [github.com/opentreeoflife/phylesystem](https://github.com/opentreeoflife/phylesystem).
  The main
  reason for this is that trees in phylesystem have a set of user defined characteristics
  that are essential for automatizing the phylogeny update process. The most relevant of these being the standardization of taxonomic names and the definition of ingroup and outgroup. Outgroup and ingroup taxa in the original tree are identified and tagged. This allows to automatically set the root for the updated tree on the final steps of the pipeline.
  Currently, only trees connected to
  a published study can be stored in the phylesystem (although there are plans to
  allow storage of unpublished trees).
  A user can choose from among the `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)). If the published tree you are interested in updating is not in there, you can upload it via OpenTree's curator tool ([www.opentreeoflife.org/curator](https://tree.opentreeoflife.org/curator).
- THE INPUT TREE CAN ALSO BE A PERSONAL ONE. A TNRS MATCHING NEEDS TO BE PERFORMED (AND A FILE GENERATED) IN ORDER TO UPDATE A PERSONAL UNPUBLISHED TREE.
- The alignment should be a gene alignment that was used to generate the tree. The original
  alignments are usually stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or the journal were the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper can download it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- A taxon name matching step is performed to verify that all taxon names on the tips
  of the tree are in the DNA character matrix and vice versa.
- A ".csv" file with the summary of taxon name matching is produced for the user.
- Unmatched taxon names are dropped from both the tree and alignment.
  Technically, just one matching name is needed to perform the searches. Please, see next section.
- A ".tre" file and a ".fas" file containing only the matched taxa are generated and saved in the `inputs` folder to be used in the following steps.

@@

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
- The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local-database sequence searches.
  For remote-database searches, we modified the BioPython [@cock2009biopython] BLAST function in the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternate BLAST URL. This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST without NCBI's required wait times.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- A pairwise BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accession numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the "physcraper" folder, allowing for secondary analyses to run significantly faster.
- Matched sequences will be discarded if they fall below a default e-value of 0.00001, and outside a default minimum and maximum length of 80% and 120%, respectively, of the average length (gaps dropped) of sequences in the checked alignment .
These parameters can be configured for each run.
  This filtering guarantees the exclusion of whole genome sequences. EXPLAIN WHY THIS IS IMPORTANT.
  All accepted sequences are assigned an internal identifier, and are further filtered.

- New sequences that are identical to existing sequences, or to subsets of an existing sequence are discarded, unless they reperesent a different taxon than the existing sequence.
Longer sequences belonging to the
  same taxon as the original sequence will be considered further for analysis.
- Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
- Reverse complement sequences are identified and translated.
- Users can choose to perform cycles of sequence similarity search iteratively, by blasting the newly found sequences until no new sequences are found. By default only one BLAST cycle is performed and only sequences in the checked alignment are blasted.
- Accepted sequences are then downloaded in full, and stored as a local database in a directory that is globally accessible (default to "physcraper/taxonomy" folder), so they are accessible for further runs.
- A fasta file containing all filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

- The software MUSCLE [@edgar2004muscle] is used by default to perform sequence alignments.
- First, all new sequences are aligned using default MUSCLE options.
- Then, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
- The final alignment is not further processed automatically. So, we encourage users
  to check it either by eye and perform manual refinement or using any of the many
  tools for alignment processing, to eliminate columns with no information.
- Users may also use Physcraper to only gather matched sequences, and apply their own preferred alignment and phylogenetic inference methods.

@@

- A gene tree is reconstructed for each alignment provided, using a Maximum Likelihood approach implemented with the software RAxML [@stamatakis2014raxml], using default settings such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with default method. Currently only the number of bootsrap replicates can be modified by the user.
- The original tree is used as starting tree for the ML searches. It can also be set as a full topological constraint or not used at all, depending on the goals of the user.
- Bootstrap results are summarized with the SumTrees module of DendroPy [current version 4.4.0; @sukumaran2010dendropy].
- The final result is an updated phylogenetic hypothesis for each of the genes provided in the alignment.
- Tips on all trees generated by Physcraper are defined by a taxon name space, capturing the NCBI accession information, as well as the taxon identifiers, allowing the user perform comparisons and conflict analyses.
- Robinson Foulds weighted and unweighted metrics between the tips in the input tree and those tips in the updated tree are calculated with Dendropy functions [@sukumaran2010dendropy].
- Finally a conflict analysis is performed. This is basically a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
- For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined.
- A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroups so trees are accurately rooted.

@@

We will illustrate the utility of Physcraper in here with two use-case scenarios. One in which the user is interested in a particular group. Another one in which the user is interested in a particular phylogeny.
A tutorial as well as illustrated examples of commands for every step needed to perform a Physcraper analysis are available elsewhere.

@@

A student is interested in the genus *Ilex*, the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.
The genus encompasses between 400-600 living species. A review of literature reveals that there are three published phylogenetic trees showing relationships within the hollies.
The first one has been made available in TreeBASE as well as in the OpenTree phylesystem and is part of the synthetic tree. It samples 48 species.
The second tree has not been made available anywhere, not even in the supplementary data of the original publication.
The most recent one has been made available in the OpenTree Phylesystem and in the DRYAD repository. It is the best sampled yet, with 200 species. However, it has not been added to the syntehtic tree yet.
This also makes it a perfect case to test the basic functionalities of Physcraper: we know that the sequences of the most recently published tree have been made available on the GenBank database. Hence, we expect that updating the oldest tree will produce something very similar to the newest tree.

@@

All these tools are key efforts for advancing towards reproducibility in phylogenetics,
a field that has relied on processes which are somewhat artisanal.
In here, we highlight the potential of taking advantage of this careful curation work in previous phylogenetic estimates. By taking sources of information available from data repositories and present a method to link data from different repositories, while leveraging on the knowledge and intuition of the expert
community to build up our phylogenetic knowledge, we can use not only data accumulated in
molecular data repositories, but phylogenetic knowledge accumulated in phylogenetic tree repositories.
While not generating full phylogenies *de novo*, Physcraper is still capable of generating new phylogenetic knowledge.
It can also be combined with data from repositories other than molecular data. For example geographic locations (GBIF), fossils (PBDB), etc.
